Parameterization of the responses of subarctic European vegetation to key environmental variables for ozone risk assessment

### 2022-04-13 VegOzone.tex
Harmonize se of key environmental variables

\begin{abstract}
  The unique vegetation of the subarctic region acclimatized to extremes of cold and midnight sun are likely to be at threat from the combined impacts of climate change and increasing ozone concentrations \chem{[O_3]}. The atmospheric and climatic characteristics of the subarctic are known to lead to pronounced peak \chem{[O_3]} in spring. To date, only a few studies assessed the response of subarctic vegetation to variations in climate and air pollution. This study looks to fill this knowledge gap by examining the uncertainty in risk assessments that are using more general boreal paramterizations. With no bespoke parameterization of stomatal response of subarctic vegetation to essential climate variables available, we propose an adjustment of existing, Boreal parameterizations, based on local climate condition. To this end, we vary temperature response functions according to the climatological temperature distribution within the growing season at our site. We evaluate the extent to which two recent years (2018 and 2019) deviate from climatic and \chem{[O_3]} norms and how these potentially more frequent future deviations may influence the projected ozone damage to subarctic vegetation. We find that 2018 was an anomalously warm and less cloudy year, particularly in spring and early summer. Higher than average \chem{[O_3]} was observed in April/May while frequent episodes of ozone volume mixing ratios (VMRs) above $40\,\unit{ppb}$ occurred in June--August. These episodes are in part attributable to forest fires in the Northern Hemisphere and warmer and sunnier conditions. We apply the integrated flux-metric Phytotoxic Ozone Dose (POD) to determine ozone risk and damage to vegetation as a function of \chem{[O_3]}, environmental factors, and species-specific physiology. Our study suggests that using standard boreal parameterizations in assessments likely leads to underestimating the risk of ozone damage in this region. We find that local climate adjusted parameterizations of plant functional types (PFTs) for subarctic vegetation result in an ozone-induced biomass loss of $2.5$ to $17.4\,\unit{\%}$. For some species, this loss is up to $xx\,\unit{\%}$ larger than projected from standard boreal parameterizations used in general assessments. Efforts should be targeted towards accurately defining subarctic vegetations' physiological response to essential climate variables.
\end{abstract}

@@

Ozone acts as oxidative stress to plants. Its main action is imposed through reactions occurring in the cell walls and cell membranes of mesophyll cells inside the leaves. Ozone enters the leaves through stomata, leaf pores that enable gas exchange, allowing the entry of \chem{CO_2} for photosynthesis and loss of \chem{H_2O} vapor via the plant transpiration stream. Stomatal aperture, and hence stomatal conductance to these gases (including ozone) will vary over the day and growing season primarily to balance \chem{CO_2} uptake against \chem{H_2O} vapor loss. The higher the stomatal conductance, the higher the potential for ozone uptake. Stomatal conductance has been empirically linked to environmental factors such as air temperature $T$, photosynthetic photon flux density (PPFD), vapor pressure deficit (VPD), and soil water potential (SWP) as well as photosynthesis itself \citep[e.g.,][]{PTRS:Jarvis1976, BallBerry1987, Emberson2000, ICP:MappingManual2017}. Results from open-top chamber (OTC) experiments on downy birch (\emph{Betula pubescens}) and mountain birch (\emph{Betula pubescens toruosa}), native to subarctic regions, indicated reductions in both biomass, root:shoot ratio, and visible leaf damage under elevated ozone treatment ($\left<\chem{[O_3]}\right> = 36-54\,\unit{ppb}$) \citep{Amb:Manninen2009}. Though, Scots pine (\emph{Pinus sylvestris}) is considered to be more ozone tolerant due to an absence of visible injuries \citep{Amb:Girgzdiene2009}, \citet{Amb:Manninen2009} found chlorophyll:carotenoid ratio and polyamines reductions under elevated ozone concentrations indicating susceptibility to ozone also in these species.\\

To estimate the potential ozone risk, we use a flux-based metric, also referred to as the Phytotoxic Ozone Dose over a threshold y ($\mathrm{POD_y}$). Flux-based metrics, that relate the accumulated dose of ozone to vegetation damage, are biologically more relevant for risk assessments than exposure-based metrics because they can account for particular species characteristics (i.e. physiology and phenology) as well as environmental conditions that can decouple the relationship between ozone concentration, ozone uptake, and consequent damage \citep{PT:Emberson2020}. These flux-based metrics are better suited to represent the actual risk to vegetation from ozone, especially in the climatically extreme parts of Europe \citep{EP:Simpson2007,GCB:Mills2011,ICP:MappingManual2017}. But no bespoke parameterization for subarctic vegetation is available.
An estimate of the stomatal \chem{O_3} flux ($\Phi_\mathrm{sto}$) is calculated based on the assumption that the concentration of \chem{O_3} at the top of the canopy $h$ represents a reasonable estimate of the concentration at the upper surface of the laminar layer for a sunlit upper canopy leaf and environmental factors affecting stomatal conductance. $\mathrm{POD_y}$ is then calculated according to:
%
\begin{equation}
  \mathrm{POD_y} = \int{(\Phi_\mathrm{sto}-y)\cdot \mathrm{d}t},
  \label{eq:pod}
\end{equation}
%
with the hourly averaged stomatal ozone flux $\Phi_\mathrm{sto}$ (see Eq.~(\ref{eq:flux_stomata})) and a stomatal ozone flux threshold $\mathrm{y}$ both given in units of \unit{nmol\,\chem{O_3}\,m^{-2}\,\mathrm{PLA}^{-1}\,s^{-1}}. Herein PLA stands for Projected~Leaf~Area. The timestep $\mathrm{d}t$ is usually 1~hour. The flux threshold~$\mathrm{y}$ represents the detoxification potential of the plant and is typically only exceeded during daylight hours (i.e. when global radiation is above $50\,\unit{W\,m^{-2}}$). $\mathrm{POD_y}$ is given in \unit{mmol\,m^{-2}} integrated over the growing season.

@@

\subsubsection{Model description}
\label{subsec:model_describtion}
The opening of stomata is governed by different species-specific and environmental factors. In $\mathrm{DO_3SE}$ it is described by a multiplicative empirical model \citep{PTRS:Jarvis1976,Emberson2000,ICP:MappingManual2017}:
\begin{equation}
  g^k_\text{sto} = g^k_\text{max} \cdot f^k_\text{phen} \cdot f^k_\text{light} \cdot \max{\left\{f^k_\text{min}, f^k_\mathrm{T} \cdot f^k_\mathrm{VPD} \cdot f^k_\text{SWP}\right\}}.
  \label{eq:stomatal}
\end{equation}
%
Where $g_\mathrm{max}$ is a species-specific maximum stomatal conductance value that is then modified by seasonal and environmental factors that vary within a range $0-1$. These factors are empirically determined and account for leaf phenology ($f_\text{phen}$), light ($f_\text{light}$), temperature ($f_\mathrm{T}$), water vapor pressure deficit ($f_\mathrm{VPD}$), and soil water potential ($f_\text{SWP}$). All factors differ with plant functional type (PFT) denoted with $k$.

@@

As indicated above, a better acclimation to prevailing climate conditions is characterized by a higher average $\left<g_\mathrm{sto}\right>/g_\mathrm{max}$. To compare our bespoke parameterizations, we propose the following metric. We calculate an average relative stomatal conductance $\left<g_\mathrm{sto}\right>/g_\mathrm{max}$ (Eq.~(\ref{eq:stomatal}) with $f_\mathrm{SWP}=1$) at noon ($11\,\unit{am}-1\,\unit{pm}$ local time) and in the morning ($5-9\,\unit{am}$ local time) for the whole record of key environmental factors' data and apply mean and standard deviation. We presume that $\left<g_\mathrm{sto}\right>/g_\mathrm{max}$ around noon (highest light intensity) is a good proxy for \chem{CO_2} uptake efficiency (growth). In addition, a small deviation between noon and morning and a low standard deviation would indicate higher robustness to variability in growing conditions. Based on these criteria, we identify \emph{subarctic}--PPFD0.8 as the best parameterization for all PFTs (Fig.~\ref{fig:javis_func_opt_mean}). As expected due to the small adjustments, coniferous trees display the smallest differences between the different parameterizations, while the differences for perennial grassland are substantial as a response to the proposed temperature acclimation.

@@

For our study, we conducted a similar bio-monitoring study at Svanhovd, a site with agrometeorological and air pollution monitoring, to assess whether local ozone concentrations were capable of inducing damage to vegetation growing under northern Fennoscandia conditions. In 2018, we indeed observed ozone damage on clover and tobacco. In contrast, no such damage was found on the clovers in 2019. This indicates that environmental factors in 2018 differed from 2019.

>>>>

\begin{abstract}
  The unique vegetation of the subarctic region acclimatized to extremes of cold and midnight sun are likely to be at threat from the combined impacts of climate change and increasing ozone concentrations \chem{[O_3]}. The atmospheric and climatic characteristics of the subarctic are known to lead to pronounced peak \chem{[O_3]} in spring. To date, only a few studies assessed the response of subarctic vegetation to variations in climate and air pollution. This study looks to fill this knowledge gap by examining the uncertainty in risk assessments that are using more general boreal paramterizations. With no bespoke parameterization of stomatal response of subarctic vegetation to essential climate variables available, we propose an adjustment of existing, Boreal parameterizations, based on local climate condition. To this end, we vary temperature response functions according to the climatological temperature distribution within the growing season at our site. We evaluate the extent to which two recent years (2018 and 2019) deviate from climatic and \chem{[O_3]} norms and how these potentially more frequent future deviations may influence the projected ozone damage to subarctic vegetation. We find that 2018 was an anomalously warm and less cloudy year, particularly in spring and early summer. Higher than average \chem{[O_3]} was observed in April/May while frequent episodes of ozone volume mixing ratios (VMRs) above $40\,\unit{ppb}$ occurred in June--August. These episodes are in part attributable to forest fires in the Northern Hemisphere and warmer and sunnier conditions. We apply the integrated flux-metric Phytotoxic Ozone Dose (POD) to determine ozone risk and damage to vegetation as a function of \chem{[O_3]}, key environmental variables, and species-specific physiology. Our study suggests that using standard boreal parameterizations in assessments likely leads to underestimating the risk of ozone damage in this region. We find that local climate adjusted parameterizations of plant functional types (PFTs) for subarctic vegetation result in an ozone-induced biomass loss of $2.5$ to $17.4\,\unit{\%}$. For some species, this loss is up to $xx\,\unit{\%}$ larger than projected from standard boreal parameterizations used in general assessments. Efforts should be targeted towards accurately defining subarctic vegetations' physiological response to essential climate variables.
\end{abstract}

@@

Ozone acts as oxidative stress to plants. Its main action is imposed through reactions occurring in the cell walls and cell membranes of mesophyll cells inside the leaves. Ozone enters the leaves through stomata, leaf pores that enable gas exchange, allowing the entry of \chem{CO_2} for photosynthesis and loss of \chem{H_2O} vapor via the plant transpiration stream. Stomatal aperture, and hence stomatal conductance to these gases (including ozone) will vary over the day and growing season primarily to balance \chem{CO_2} uptake against \chem{H_2O} vapor loss. The higher the stomatal conductance, the higher the potential for ozone uptake. Stomatal conductance has been empirically linked to key environmental variables such as air temperature $T$, photosynthetic photon flux density (PPFD), vapor pressure deficit (VPD), and soil water potential (SWP) as well as photosynthesis itself \citep[e.g.,][]{PTRS:Jarvis1976, BallBerry1987, Emberson2000, ICP:MappingManual2017}. Results from open-top chamber (OTC) experiments on downy birch (\emph{Betula pubescens}) and mountain birch (\emph{Betula pubescens toruosa}), native to subarctic regions, indicated reductions in both biomass, root:shoot ratio, and visible leaf damage under elevated ozone treatment ($\left<\chem{[O_3]}\right> = 36-54\,\unit{ppb}$) \citep{Amb:Manninen2009}. Though, Scots pine (\emph{Pinus sylvestris}) is considered to be more ozone tolerant due to an absence of visible injuries \citep{Amb:Girgzdiene2009}, \citet{Amb:Manninen2009} found chlorophyll:carotenoid ratio and polyamines reductions under elevated ozone concentrations indicating susceptibility to ozone also in these species.\\

To estimate the potential ozone risk, we use a flux-based metric, also referred to as the Phytotoxic Ozone Dose over a threshold y ($\mathrm{POD_y}$). Flux-based metrics, that relate the accumulated dose of ozone to vegetation damage, are biologically more relevant for risk assessments than exposure-based metrics because they can account for particular species characteristics (i.e. physiology and phenology) as well as key environmental variables that can decouple the relationship between ozone concentration, ozone uptake, and consequent damage \citep{PT:Emberson2020}. These flux-based metrics are better suited to represent the actual risk to vegetation from ozone, especially in the climatically extreme parts of Europe \citep{EP:Simpson2007,GCB:Mills2011,ICP:MappingManual2017}. But no bespoke parameterization for subarctic vegetation is available.
An estimate of the stomatal \chem{O_3} flux ($\Phi_\mathrm{sto}$) is calculated based on the assumption that the concentration of \chem{O_3} at the top of the canopy $h$ represents a reasonable estimate of the concentration at the upper surface of the laminar layer for a sunlit upper canopy leaf and key environmental variables affecting stomatal conductance. $\mathrm{POD_y}$ is then calculated according to:
%
\begin{equation}
  \mathrm{POD_y} = \int{(\Phi_\mathrm{sto}-y)\cdot \mathrm{d}t},
  \label{eq:pod}
\end{equation}
%
with the hourly averaged stomatal ozone flux $\Phi_\mathrm{sto}$ (see Eq.~(\ref{eq:flux_stomata})) and a stomatal ozone flux threshold $\mathrm{y}$ both given in units of \unit{nmol\,\chem{O_3}\,m^{-2}\,\mathrm{PLA}^{-1}\,s^{-1}}. Herein PLA stands for Projected~Leaf~Area. The timestep $\mathrm{d}t$ is usually 1~hour. The flux threshold~$\mathrm{y}$ represents the detoxification potential of the plant and is typically only exceeded during daylight hours (i.e. when global radiation is above $50\,\unit{W\,m^{-2}}$). $\mathrm{POD_y}$ is given in \unit{mmol\,m^{-2}} integrated over the growing season.

@@

\subsubsection{Model description}
\label{subsec:model_describtion}
The opening of stomata is governed by different species-specific factors and response to key environmental variables. In $\mathrm{DO_3SE}$ it is described by a multiplicative empirical model \citep{PTRS:Jarvis1976,Emberson2000,ICP:MappingManual2017}:
\begin{equation}
  g^k_\text{sto} = g^k_\text{max} \cdot f^k_\text{phen} \cdot f^k_\text{light} \cdot \max{\left\{f^k_\text{min}, f^k_\mathrm{T} \cdot f^k_\mathrm{VPD} \cdot f^k_\text{SWP}\right\}}.
  \label{eq:stomatal}
\end{equation}
%
Where $g_\mathrm{max}$ is a species-specific maximum stomatal conductance value that is then modified by seasonal and response functions to key environmental variables that vary within a range $0-1$. These factors are empirically determined and account for leaf phenology ($f_\text{phen}$), light ($f_\text{light}$), temperature ($f_\mathrm{T}$), water vapor pressure deficit ($f_\mathrm{VPD}$), and soil water potential ($f_\text{SWP}$). All factors differ with plant functional type (PFT) denoted with $k$.

@@

As indicated above, a better acclimation to prevailing climate conditions is characterized by a higher average $\left<g_\mathrm{sto}\right>/g_\mathrm{max}$. To compare our bespoke parameterizations, we propose the following metric. We calculate an average relative stomatal conductance $\left<g_\mathrm{sto}\right>/g_\mathrm{max}$ (Eq.~(\ref{eq:stomatal}) with $f_\mathrm{SWP}=1$) at noon ($11\,\unit{am}-1\,\unit{pm}$ local time) and in the morning ($5-9\,\unit{am}$ local time) for the whole record of key environmental variables' data and apply mean and standard deviation. We presume that $\left<g_\mathrm{sto}\right>/g_\mathrm{max}$ around noon (highest light intensity) is a good proxy for \chem{CO_2} uptake efficiency (growth). In addition, a small deviation between noon and morning and a low standard deviation would indicate higher robustness to variability in growing conditions. Based on these criteria, we identify \emph{subarctic}--PPFD0.8 as the best parameterization for all PFTs (Fig.~\ref{fig:javis_func_opt_mean}). As expected due to the small adjustments, coniferous trees display the smallest differences between the different parameterizations, while the differences for perennial grassland are substantial as a response to the proposed temperature acclimation.

@@

For our study, we conducted a similar bio-monitoring study at Svanhovd, a site with agrometeorological and air pollution monitoring, to assess whether local ozone concentrations were capable of inducing damage to vegetation growing under northern Fennoscandia conditions. In 2018, we indeed observed ozone damage on clover and tobacco. In contrast, no such damage was found on the clovers in 2019. This indicates that key environmental variables in 2018 differed from 2019.
